---
title: "Strandedness of promotors"
output: html_notebook
author: Adrianna Vandeuren
date: 10/31/2024
---

The goal of this script is to assign strandedness to promotors and TssA regions derived from the roadmap epigenomics consortium data based on the genes they control. 

Here is the code for promotors. The rationale behind this code is:
1) Load all packages and required data
2) Create a list of start locations of genes based on their strandedness (meaning: genes on + strand start = start; genes on - strand start = end).
3) Based on the accurate start location list, look for the nearest start location to the promotors and assign the associated strandedness to the promotor


```{r}
## Load required libraries
library(GenomicRanges)
library(rtracklayer)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)

## Set the working directory to where your feature files are
setwd("~/path/to/feature/of/interest/directory")

## Define the path to your BED file
promoter_file_path <- "~/path/to/promotor/file.bed"

# Import promoter BED file
promoters <- import(promoter_file_path)

# Load hg19 gene annotation from TxDb
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

# Extract genes
genes <- genes(txdb)

# Create a GRanges object for adjusted gene starts based on strandedness
adjusted_gene_starts <- GRanges(seqnames = seqnames(genes),
                                ranges = IRanges(start = ifelse(strand(genes) == "+", 
                                                                start(genes), 
                                                                end(genes)),  # Use start for + strand, end for - strand
                                                end = ifelse(strand(genes) == "+", 
                                                             start(genes), 
                                                             end(genes))),
                                strand = strand(genes))

# Initialize vector to store nearest gene indices and their strands
nearest_gene_indices <- vector("integer", length(promoters))
nearest_strands <- vector("character", length(promoters))

# Find the nearest gene for each promoter based on adjusted start locations
for (i in seq_len(length(promoters))) {
  promoter <- promoters[i]
  
  # Find the nearest adjusted gene start
  nearest_gene_indices[i] <- nearest(promoter, adjusted_gene_starts)
  
  # Assign the strand of the nearest gene
  nearest_strands[i] <- as.character(strand(adjusted_gene_starts[nearest_gene_indices[i]]))
}

# Assign the strand information to the promoters
promoters_with_strand <- promoters
strand(promoters_with_strand) <- nearest_strands

# Check if the strand information is set correctly
# Print first few promoters to confirm
print(head(promoters_with_strand))

# Export the updated promoters with assigned strand information
export(promoters_with_strand, "promoters_with_assigned_strand.bed", format = "BED")
```
Separate the resulting files based on strand:
```{r}
# Load necessary libraries
library(GenomicRanges)
library(rtracklayer)

# Assume 'promoters_with_strand' is your GRanges object with strand information already assigned

# Subset promoters on the positive strand
promoters_plus <- promoters_with_strand[strand(promoters_with_strand) == "+"]
# Subset promoters on the negative strand
promoters_minus <- promoters_with_strand[strand(promoters_with_strand) == "-"]

# Export the subsets to separate BED files
export(promoters_plus, "promoters_with_G4Peaks_plus_strand.bed", format = "BED")
export(promoters_minus, "promoters_with_G4Peaks_minus_strand.bed", format = "BED")
```
